Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-08-27 LICS/sections/appendix.tex
added missing github action

\begin{theorem}[Existence of pushouts in $\catname{EHyp_{\Sigma}}$]

@@

Consider the following span in $\catname{EHyp_{\Sigma}}$

@@

We then quotient the set vertices and edges in $X + Y$ by $R_{V}$ and $R_{E}$ respectively.
\ifdefined \ONECOLUMN
\begin{align*}
    \sfrac{X + Y}{\sim (R_{V},R_{E})} = \{
        &\sfrac{V_{X} + V_{Y}}{\sim (R_{V},R_{E})}, \sfrac{E_{X} + E_{Y}}{\sim (R_{V},R_{E})}, \sfrac{s_{X+Y}}{\sim (R_{V},R_{E})},\\
        &\sfrac{t_{X,Y}}{\sim (R_{V},R_{E})}, \consistency_{\sfrac{X+Y}{\sim (R_{V},R_{E})}}, <_{\sfrac{X + Y}{\sim (R_{V},R_{E})}}\}    
\end{align*}
\else
    \begin{align*}
        \sfrac{X + Y}{\sim (R_{V},R_{E})} = \{&\\
            &\sfrac{V_{X} + V_{Y}}{\sim (R_{V},R_{E})}, \sfrac{E_{X} + E_{Y}}{\sim (R_{V},R_{E})}, \sfrac{s_{X+Y}}{\sim (R_{V},R_{E})},\\
            &\sfrac{t_{X,Y}}{\sim (R_{V},R_{E})}, \consistency_{\sfrac{X+Y}{\sim (R_{V},R_{E})}}, <_{\sfrac{X + Y}{\sim (R_{V},R_{E})}}\\
        & \}    
    \end{align*}
\fi
We will refer to $\sim (R_{V},R_{E})$ just as $\sim$, \textit{e.g.}, by writing $\sfrac{V_{X} + V_{Y}}{\sim}$ and concrete relation will be clear from the context.
Where necessary we will refer to $S_{V}$ and $S_{R}$ as $\sim_{S}$.
We have 
\[
    \sfrac{s_{X+Y}}{\sim} : \sfrac{E_{X} + E_{Y}}{\sim} \to (\sfrac{V_{X} + V_{Y}}{\sim})^{*}
\]
and there is an obvious surjective function $[-]_{V} : (V_{X} + V_{Y}) \to (\sfrac{V_{X} + V_{Y}}{\sim})$ that maps elements to their equivalence classes and $[-]_{V}^{*}$ is its extension to sequences.
There is also $[-] : E_{X} + E_{Y} \to \sfrac{E_{X} + E_{Y}}{\sim}$ and we will omit subscripts as the correct type will be clear from the argument.
We then define 
\[
    \sfrac{s_{X+Y}}{\sim}([e]) = s_{X+Y};[-]^{*}(e) = [s_{X+Y}(e)]^{*}
\]
We will also use subscripts when it is important to tell if an element of $E_{X} + E_{Y}$ has a pre-image in either $E_{X}$ or $E_{Y}$ by writing $e_{x}$ or $e_{y}$.
Similarly, we will write $v_{x}$ to refer to a vertex with a pre-image in $V_{X}$.
Likewise, 
\[
    t_{\sfrac{X+Y}{\sim}}([e]) = [t_{X+Y}(e)]^{*}
\]

@@

% So far, the construction matches the construction of the pushout in $\catname{Hyp_{\Sigma}}$ which can be found in~\cite{bonchi_string_2022-1} and hence $s_{\sfrac{X+Y}{\sim}}$ and $t_{\sfrac{X+Y}{\sim}}$ are well-defined and $[-]$ is a homomorphism with respect to sources and targets and therefore $\sfrac{\iota_1}{\sim}$ and $\sfrac{\iota_{2}}{\sim}$ are.

@@

\begin{figure}
    \[
        \scalebox{0.5}{
        \tikzfig{../figures/appendix/pushout_example}
        }
\]
\captionsetup{belowskip=-3ex}
\caption{Pushout in $\catname{EHyp_{\Sigma}}$}
\label{fig:pushout_example}
\end{figure}

@@

We will then discuss the uniqueness of the corresponding pushout complement.

@@

\begin{proposition}

@@

        The pushout in~\ref{def:boundary_new} satisfies the assumptions in Definition~\ref{pushout:assumptions} (note that since $i \to i' \to \mathcal{L} \xleftarrow{} j' \xleftarrow{} j$ is an MDA cospan the image of $i + j$ in $\mathcal{L}$ consists of top-level vertices only) and therefore is isomorphic to the pushout constructed in~\ref{pushout:assumptions} and by construction should also be a pushout on the underlying sets of vertices and hyperedges.

@@

@@

            \] is an MDA e-hypergraph.

@@

            one can observe that $\llbracket f \rrbracket$ is the pushout object of $\mathcal{L} \xleftarrow{} i+j \xrightarrow{} \mathcal{L}$ as per the diagram below

@@

    The rest of the argument follows by induction on the length of $w$ and ultimately we have shown that if $f = g$ then $\llbracket f \rrbracket = \llbracket g \rrbracket$ in $\WellTypedMdaEcospans(\Sigma)/\mathcal{S}$.

@@

    % As SMC laws are subsumed by e-hypergraph representation we need to show that for each enrichment equation $l = r$ that turns $g$ into $h$ there is a pushout complement $\mathcal{C}$ such that $\llbracket g \rrbracket = \mathcal{C}[\llbracket l \rrbracket]$ and such that $\llbracket h \rrbracket = \mathcal{C}[\llbracket r \rrbracket]$, where $\mathcal{C}[-]$ is defined by computing the pushout along the common interfaces.

@@

>>>>

\begin{theorem}[Existence of pushouts in $\catname{EHyp}(\Sigma)$]

@@

Consider the following span in $\catname{EHyp}(\Sigma)$

@@

We then quotient the set vertices and edges in $X + Y$ by $R_{V}$ and $R_{E}$ respectively.
\ifdefined \ONECOLUMN
\begin{align*}
    \sfrac{X + Y}{\sim (R_{V},R_{E})} = \{
        &\sfrac{V_{X} + V_{Y}}{\sim (R_{V},R_{E})}, \sfrac{E_{X} + E_{Y}}{\sim (R_{V},R_{E})}, \sfrac{s_{X+Y}}{\sim (R_{V},R_{E})},\\
        &\sfrac{t_{X,Y}}{\sim (R_{V},R_{E})}, \consistency_{\sfrac{X+Y}{\sim (R_{V},R_{E})}}, <_{\sfrac{X + Y}{\sim (R_{V},R_{E})}}\}    
\end{align*}
\else
    \begin{align*}
        \sfrac{X + Y}{\sim (R_{V},R_{E})} = \{&\\
            &\sfrac{V_{X} + V_{Y}}{\sim (R_{V},R_{E})}, \sfrac{E_{X} + E_{Y}}{\sim (R_{V},R_{E})}, \sfrac{s_{X+Y}}{\sim (R_{V},R_{E})},\\
            &\sfrac{t_{X,Y}}{\sim (R_{V},R_{E})}, \consistency_{\sfrac{X+Y}{\sim (R_{V},R_{E})}}, <_{\sfrac{X + Y}{\sim (R_{V},R_{E})}}\\
        & \}    
    \end{align*}
\fi
We will refer to $\sim (R_{V},R_{E})$ just as $\sim$, \textit{e.g.}, by writing $\sfrac{V_{X} + V_{Y}}{\sim}$ and concrete relation will be clear from the context.
Where necessary we will refer to $S_{V}$ and $S_{R}$ as $\sim_{S}$.
We have 
\[
    \sfrac{s_{X+Y}}{\sim} : \sfrac{E_{X} + E_{Y}}{\sim} \to (\sfrac{V_{X} + V_{Y}}{\sim})^{*}
\]
and there is an obvious surjective function $[]_{V} : (V_{X} + V_{Y}) \to (\sfrac{V_{X} + V_{Y}}{\sim})$ that maps elements to their equivalence classes and $[]_{V}^{*}$ is its extension to sequences.
There is also $[] : E_{X} + E_{Y} \to \sfrac{E_{X} + E_{Y}}{\sim}$ and we will omit subscripts as the correct type will be clear from the argument.
We then define 
\[
    \sfrac{s_{X+Y}}{\sim}([e]) = s_{X+Y};[]^{*}(e) = [s_{X+Y}(e)]^{*}
\]
We will also use subscripts when it is important to tell if an element of $E_{X} + E_{Y}$ has a pre-image in either $E_{X}$ or $E_{Y}$ by writing $e_{x}$ or $e_{y}$.
Similarly, we will write $v_{x}$ to refer to a vertex with a pre-image in $V_{X}$.
Likewise, 
\[
    t_{\sfrac{X+Y}{\sim}}([e]) = [t_{X+Y}(e)]^{*}
\]

@@

% So far, the construction matches the construction of the pushout in $\catname{Hyp_{\Sigma}}$ which can be found in~\cite{bonchi_string_2022-1} and hence $s_{\sfrac{X+Y}{\sim}}$ and $t_{\sfrac{X+Y}{\sim}}$ are well-defined and $[]$ is a homomorphism with respect to sources and targets and therefore $\sfrac{\iota_1}{\sim}$ and $\sfrac{\iota_{2}}{\sim}$ are.

@@

\begin{remark}
    If in a span $Y \xleftarrow{g} Z \xrightarrow{f} X$ of morphisms in $\catname{EHyp}(\Sigma)$ the morphisms $f$ and $g$ are monos, then the arrows $f'$ and $g'$ of the pushout $Y \xrightarrow{g'} X +_{f,g} Y \xleftarrow{f'} X$ are also monos.
\end{remark}
This is because $[]_{V}$ identifies two vertices $v_1$ and $v_2$ only if there exists a vertex $z$ in $V_{Z}$ such that $f_{V};\iota_{1,V}(z) = v_1$ and $g_{V};\iota_{2,V}(z) = v_2$ and therefore $f' = \iota_{1};[]$ and $g' = \iota_{2};[]$ are monos.

\begin{figure}
    \[
        \scalebox{0.5}{
        \tikzfig{../figures/appendix/pushout_example}
        }
\]
\captionsetup{belowskip=-3ex}
\caption{Pushout in $\catname{EHyp}(\Sigma)$}
\label{fig:pushout_example}
\end{figure}

@@

We will then discuss the uniqueness of the corresponding boundary pushout complement.

@@

\begin{proposition}[Proposition~\ref{prop:boundary_unique} restatement]

@@

        The pushout in~\ref{def:boundary_new} satisfies the assumptions in Definition~\ref{pushout:assumptions} (note that since $i \to i' \to \mathcal{L} \xleftarrow{} j' \xleftarrow{} j$ is an MDA cospan, the image of $i + j$ in $\mathcal{L}$ consists of top-level vertices only) and therefore is isomorphic to the pushout constructed in~\ref{pushout:assumptions} and by construction should also be a pushout on the underlying sets of vertices and hyperedges.

@@

\begin{remark}
    Below we will also occasionally conflate the cospan resulting from $\llbracket - \rrbracket$ with its carrier which will be clear from the context.
    For example, if $\llbracket f \rrbracket = n \to n' \to \mathcal{F} \xleftarrow{} m' \xleftarrow{} m$, then we will also use $\llbracket f \rrbracket$ to denote $\mathcal{F}$.
\end{remark}

@@

            \] is an MDA cospan.

@@

            one can observe that $\llbracket f \rrbracket$ is the pushout object of $\mathcal{L} \xleftarrow{} i+j \xrightarrow{} \mathcal{L}_{3}^{\bot}$ as per the diagram below

@@

    The rest of the argument follows by induction on the length of $w$ and ultimately we have shown that if $f = g$ then $\llbracket f \rrbracket = \llbracket g \rrbracket$ in $\WellTypedMdaEcospans / \mathcal{S}$.

@@

    % As SMC laws are subsumed by e-hypergraph representation we need to show that for each enrichment equation $l = r$ that turns $g$ into $h$ there is a pushout complement $\mathcal{C}$ such that $\llbracket g \rrbracket = \mathcal{C}[\llbracket l \rrbracket]$ and such that $\llbracket h \rrbracket = \mathcal{C}[\llbracket r \rrbracket]$, where $\mathcal{C}[]$ is defined by computing the pushout along the common interfaces.

@@

\begin{lemma}
\label{lemma:decomposition}
    Let $n \xrightarrow{f} n' \xrightarrow{f'} \mathcal{G} \xleftarrow{g'} m' \xleftarrow{g} m$ be an MDA extended cospan.
    If $\mathcal{L}$ is a convex down-closed sub-e-hypergraph of $\mathcal{G}$ such that the immediate predecessors of top-level connected components of $\mathcal{L}$ in $\mathcal{G}$ are either undefined or coincide and pair-wise consistent then either
    \begin{itemize}
    \item there exists a cospan $i \to i' \to \mathcal{G}' \xleftarrow{} j' \xleftarrow{} j$ and $k \in \mathbb{N}$ such that $\mathcal{G}$ is decomposable as
    \begin{align}
        \label{fig:decomposition:1}
    \scalebox{0.6}{
        \tikzfig{../figures/appendix/G_decomposition_1}
    }
    \end{align}
    \item or, $\mathcal{G}$ is decomposable as
    \begin{align}
        \label{fig:decomposition:2}
    \scalebox{0.6}{
        \tikzfig{../figures/appendix/G_decomposition_2}
    }
    \end{align}
    \end{itemize}
    such that $\mathcal{L}$ is a convex down-closed sub-e-hypergraph of $\mathcal{G}'$ and all the cospans are MDA.
\end{lemma}
\begin{proof}
    Suppose the image of top-level components of $\mathcal{L}$ consists of top-level edges and vertices.
    Then let $\mathcal{C}_1$ be the smallest e-hypergraph consisting of the input vertices of $\mathcal{G}$ and every hyperedge $h$ (and its successors) that is not in $\mathcal{L}$ but has a path to it.
    Let $\mathcal{C}_{2}$ be the smallest e-hypergraph such that $\mathcal{G} = \mathcal{C}_{1} \cup \mathcal{L} \cup \mathcal{C}_{2}$.
    These e-hypergraphs overlap only on vertices so we define
    \begin{align*}
        i = V_{\mathcal{C}_{1}} \cap V_{\mathcal{L}}\\
        j = V_{\mathcal{C}_{2}} \cap V_{\mathcal{L}}\\
        k = (V_{\mathcal{C}_{1}} \cap V_{\mathcal{C}_{2}}) \setminus V_{\mathcal{L}}
    \end{align*}
    and let $i'$ be the union of $i$ and the vertices of $\mathcal{L}$ in the image of $f'$ and $j'$ be the union of $j$ and the vertices of $\mathcal{L}$ in the image of $g'$.
    Let $n_{1}'$ be pre-image of the co-restriction of $f'$ to $\mathcal{C}_{1}$ and $n_{2}'$ be the pre-image of the co-restriction of $f'$ to $\mathcal{C}_{2}$.
    Similarly, for $m_{1}'$ and $m_{2}'$ with regard to $g'$.
    Then there are cospans
    \begin{align*}
        n \to n_{1}' \to &\mathcal{C}_{1} \xleftarrow{} m_{1}' + k + i \xleftarrow{} k + i\\
        i \to i' \to &\mathcal{L} \xleftarrow{} j' \xleftarrow{} j\\
        k + j \to n_{2}' + k + j \to &\mathcal{C}_{2} \xleftarrow{} m_{2}' \xleftarrow{} m\\
    \end{align*}
    and $n \xrightarrow{f} n' \xrightarrow{f'} \mathcal{G} \xleftarrow{g'} m' \xleftarrow{g} m$ is the co-limit of the following diagram
    \[
        n \to n_{1}' \to \mathcal{C}_{1} \xleftarrow{} m_{1}' + k \xleftarrow{} i + k \to i' + k \to \mathcal{L} \xleftarrow{} k + j' \xleftarrow{} k + j \xrightarrow{} n_{2}' + k + j \to \mathcal{C}_{2} \xleftarrow{} m_{2}' \xleftarrow{} m
    \]
    The two spans identify precisely those nodes from $\mathcal{G}$ that occur in more than one sub-hypergraph, so this amounts to simply taking the union
    \[
    n \to n_{1}' \cup n_{2}' \cup i' \to \mathcal{C}_{1} \cup \mathcal{C}_{2} \cup \mathcal{L} \xleftarrow{} m_{1}' \cup m_{2}' \cup j' \xleftarrow{} m  = n \xrightarrow{f} n' \xrightarrow{f'} \mathcal{G} \xleftarrow{g'} m' \xleftarrow{g} m
    \]
    If we let $\mathcal{G}' = \mathcal{L}$ we get the decomposition as in~(\ref{fig:decomposition:1}).
    This follows the construction of a similar lemma in~\cite{bonchi_string_2022-1} (Lemma 24). 
    The cospans with $\mathcal{C}_{1}$ and $\mathcal{C}_{2}$ are MDAs because they are closed under successors and predecessors and the cospan with $\mathcal{L}$ is an MDA because it is convex.

    Now suppose that the image of top-level components of $\mathcal{L}$ consist of non top-level edges and vertices $e$ and $v$.
    Take a hyperedge $h$ such that $h \leq e$ and for all $e' \not = h$ such that $e' \leq e$, $h \leq e'$.
    Let $\mathcal{G}''$ be $h$ including all its sources and targets and successors.
    If $\mathcal{G} = \mathcal{G}''$ then it is decomposable as in~(\ref{fig:decomposition:2}) and let $\mathcal{G}'$ be the component that contains the image of $\mathcal{L}$ which exists by pair-wise consistency.
    All $\mathcal{G}_{i}$ (and their respective cospans) are MDA by definitions.
    Otherwise, if $\mathcal{G} \not = \mathcal{G}''$ then $\mathcal{G}''$ is a convex down-closed sub-e-hypergraph of $\mathcal{G}$ and 
    by applying the same construction as above we get~(\ref{fig:decomposition:1}).
\end{proof}

### 2024-08-27 LICS/sections/conclusion.tex
added missing github action

In the future,  we intend to investigate a number of natural extensions to mathematical treatment of (acyclic) e-graphs developed here.
Some of the extensions that can be considered are: support for `functorial boxes'~\cite{mellies_functorial_2006} (in addition to e-boxes), which have a variety of applications, including the Cartesian-closure required for the implementation of functional programming languages~\cite{ghica_hierarchical_2023}, as sketched in Fig. \ref{fig:app1}(a); the `spiders' used by the ZX-calculus to model quantum circuits~\cite{coecke_interacting_2011,ZX}, as sketched in Fig. \ref{fig:app1}(b); or trace~\cite{joyal_geometry_1991, Hasegawa-traced} which can be used to express feedback in categorical models of digital circuits~\cite{ghica_jung_2017,ghica_compositional_2023} and, indeed in conventional e-graphs to encode infinite equivalence classes, as sketched in Fig.~\ref{fig:app2}. All these application domains stand to benefit from the use of optimisation techniques, which in turn can take advantage of the appropriate e-graph technique.

>>>>

In the future,  we intend to investigate a number of natural extensions to mathematical treatment of (acyclic) e-graphs developed here.
Some of the extensions that can be considered are: support for `functorial boxes'~\cite{mellies_functorial_2006} (in addition to e-boxes), which have a variety of applications, including the Cartesian-closure required for the implementation of functional programming languages~\cite{ghica-zanassi2023string}, as sketched in Fig. \ref{fig:app1}(a); the `spiders' used by the ZX-calculus to model quantum circuits~\cite{coecke_interacting_2011,ZX}, as sketched in Fig. \ref{fig:app1}(b); or trace~\cite{joyal_geometry_1991, Hasegawa-traced} which can be used to express feedback in categorical models of digital circuits~\cite{ghica_jung_2017,ghica_compositional_2023} and, indeed in conventional e-graphs to encode infinite equivalence classes, as sketched in Fig.~\ref{fig:app2}. All these application domains stand to benefit from the use of optimisation techniques, which in turn can take advantage of the appropriate e-graph technique.

### 2024-08-27 LICS/sections/e-hypergraphs.tex
added missing github action

\begin{figure}
    \[
    \scalebox{0.6}{
        \tikzfig{../figures/combinatorial_semantics/extended_cospan_example}
    }    
    \]
    \captionsetup{belowskip=-3ex}
    \caption{Extended cospan example}
    \label{fig:e-cospan-example}
\end{figure}

@@

    \tikzfig{../figures/combinatorial_semantics/DPOI_pushout_complement}
}
\]
is a \textit{boundary complement} if
\begin{enumerate} 
    \item $m(\mathcal L)$ is a convex down-closed e-hypergraph;
    \item $[c_1,c_2]$ is mono;
    \item for all $v,w$ in $\mathcal{G}$ in the image of $i + j$,  $v$ and $w$ share the same set of parents, and either $v,w$ are top-level or else $v \consistency w$;
    \item for all $v,w$ in $\mathcal{L}^{\bot}$ in the image of $i + j$,  $v$ and $w$ share the same set of parents, and either $v,w$ are top-level or else $v \consistency w$;
    \item there exist $d_1 : n \to \mathcal{L}^\bot$ and $d_2 : k \to \mathcal{L}^\bot$ making the above triangle commute; and
    \item if the image of $\mathcal{L}$'s external interfaces under $m$ consists exclusively of top-level vertices of $\mathcal{G}$ then there exists a \textit{well-typed} extended mda-cospan
    \[
    n + j \xrightarrow{f_1 + id_{j}} n' \setminus (i' \setminus i) + j \xrightarrow{[g_1,c_2]} \mathcal{L}^{\bot} \xleftarrow{[g_2,c_1]} k' \setminus (j' \setminus j) + i \xleftarrow{f_2 + id_{i}} k + i
    \]
    \item if the image of $\mathcal{L}$'s external interfaces under $m$ consists exclusively of not top-level vertices of $\mathcal{G}$ then there exists a \textit{not necessarily well-typed} extended mda-cospan
    \[
    n \xrightarrow{f_1} n' \setminus (i' \setminus i) + j \xrightarrow{[g_1,c_2]} \mathcal{L}^{\bot} \xleftarrow{[g_2,c_1]} k' \setminus (j' \setminus j) + i \xleftarrow{f_2} k
    \]
\end{enumerate}
where $f_i$ and $g_i$ are defined as follows.  
 %because the image of $m$ is a convex down-closed e-hypergraph, 
The strictly internal interface $i' \setminus i$ of $\mathcal L$ is mapped to the internal interface $n'$ of $\mathcal G$,  since $\mathcal L$ is a down-closed subgraph of $\mathcal G$,  inducing an identification of $i' \setminus i$ in  $n'$. 
Then map $f_1$ is given by $g_{ext}$ with its codomain restricted to $n' \setminus (i' \setminus i)$.\footnote{Noting the image of $g_{ext}$ indeed lies within $n' \setminus (i' \setminus i)$}  The map $g_1$ is derived from the restriction of $g_{int}$ to type $n' \setminus (i' \setminus i) \to \mathcal G$ by further observing that $\mathcal L^\bot$ can be identified within $\mathcal G$ --- and in particular has the internal interface $n'$ of $\mathcal G$ minus $i' \setminus i$.  The maps $f_2$ and $g_2$ are defined similarly. 
%$i' \setminus i$, which is a strict internal interface of $\mathcal{L}$, is necessarily mapped to vertices in the strict internal interface of $\mathcal{G}$ which are also in the image of $n'$ and therefore we can construct a 
%morphism $f' : i' \setminus i \to n'$ by first following $m$ and then composing with the partial inverse of $g_{int}$ which exists because it is mono. 
%Then we define $n' \setminus (i' \setminus i)$ as the pushout complement in the diagram below
%
%\[
%    \tikzfig{combinatorial_semantics/pushout_difference}
%\]
%and $g_1$ is obtained by composition $g';g_{int}$ and then composing with the inverse of $l$.
%Similarly for $g_2$.
% CHRIS: TRY TO DEFINE G1, G2 AS SIMPLY AS POSSIBLE, AND POSSIBLY ADD AN INFORMAL EXPLANATION BELOW THE DEFINITION. 
% Because $\mathcal{L}^{\bot}$ is essentially $\mathcal{G}$ with $\mathcal{L}$ removed, apart from the part that has a pre-image in $i + j$, there is an obvious morphism from $l \to \mathcal{L}^{\bot}$ which is $g_1 = g_{int} |_{D}$ where
% \[
%     D = \{ v \in n' \text{ such that } \not \exists u \in (i' \setminus i) ~ . ~ g_{int}(v) = f'_{int};f(u) \}
% \]
% Similarly for $r \to \mathcal{L}^{\bot}$, which we denote as $g_2$.
% Then there exists a morphism $n \to l$, because there is a morphism $n \to n'$ and the external interfaces of $n'$ and $l$ are the same.
% That is, it is a morphism $f: n \to l$ such that $d_1 = f;g_1$.
% Similarly for $k \to r$.
\end{definition}

Intuitively the above definition means that when the occurrence of $\mathcal{L}$ within $\mathcal{G}$ is top-level then $\mathcal{L}$'s external interfaces become the part of $\mathcal{L}^{\bot}$'s external interfaces and when the occurrence is nested, those interfaces become the part of $\mathcal{L}^{\bot}$'s strict internal interfaces.
In both cases when removing the occurrence of $\mathcal{L}$ from $\mathcal{G}$, $\mathcal{L}$'s strict internal interfaces are removed from strict internal interfaces of $\mathcal{G}$ to form internal interfaces of the corresponding mda-cospans above.
The morphisms are then constructed correspondingly to require that inputs (respectively, outputs) of $\mathcal{R}$ are glued to the outputs (respectively, inputs) of $\mathcal{L}^{\bot}$.

@@

Note that,  in the above definition,  the extended cospan 
    \[
    n \xrightarrow{} n' \setminus (i' \setminus i) + j \xrightarrow{[g_1,c_2]} \mathcal{L}^{\bot} \xleftarrow{[g_2,c_1]} k' \setminus (j' \setminus j) + i \xleftarrow{} k
    \]
 is not necessarily well-typed.
This is because its input internal interface may contain vertices that previously were a part of the output internal interface and vice versa.
\begin{remark}
The boundary complement conditions, in particular, prohibit finding a match for $\mathcal L$ in $\mathcal G$,  below.  
\[
	\mathcal{L} = \scalebox{0.8}{\tikzfig{../figures/combinatorial_semantics/f_times_g}} \qquad \mathcal{G} = \scalebox{0.6}{\tikzfig{../figures/combinatorial_semantics/f_plus_g_inline}}
\]
\end{remark}
%The uniqueness and existence properties of boundary complements as defined above are discussed in the appendix (Proposition~\ref{prop:boundary_complement} and Proposition~\ref{prop:complement_existence}).

% \begin{proposition}[Proposition~\ref{prop:boundary_complement} restatement]
%     The boundary complement in~\ref{def:boundary_new} when exists is unique.
% \end{proposition}

>>>>

\begin{figure}
    \[
    \scalebox{0.5}{
        \tikzfig{../figures/combinatorial_semantics/extended_cospan_example}
    }    
    \]
    \captionsetup{belowskip=-3ex}
    \caption{Extended cospan example}
    \label{fig:e-cospan-example}
\end{figure}

@@

    \tikzfig{../figures/combinatorial_semantics/DPOI_pushout_complement}
}
\]
is a \textit{boundary complement} if
\begin{enumerate} 
    \item $m(\mathcal L)$ is a convex down-closed e-hypergraph;
    \item $[c_1,c_2]$ is mono;
    \item for all $v,w$ in $\mathcal{G}$ in the image of $i + j$,  $v$ and $w$ share the same set of parents, and either $v,w$ are top-level or else $v \consistency w$;
    \item for all $v,w$ in $\mathcal{L}^{\bot}$ in the image of $i + j$,  $v$ and $w$ share the same set of parents, and either $v,w$ are top-level or else $v \consistency w$;
    \item there exist $d_1 : n \to \mathcal{L}^\bot$ and $d_2 : k \to \mathcal{L}^\bot$ making the above triangle commute; and
    \item if the image of $\mathcal{L}$'s external interfaces under $m$ consists exclusively of top-level vertices of $\mathcal{G}$ then there exists a \textit{well-typed} extended MDA cospan
    \[
    n + j \xrightarrow{f_1 + id_{j}} n' \setminus (i' \setminus i) + j \xrightarrow{[g_1,c_2]} \mathcal{L}^{\bot} \xleftarrow{[g_2,c_1]} k' \setminus (j' \setminus j) + i \xleftarrow{f_2 + id_{i}} k + i
    \]
    \item if the image of $\mathcal{L}$'s external interfaces under $m$ consists exclusively of not top-level vertices of $\mathcal{G}$ then there exists a \textit{not necessarily well-typed} extended MDA cospan
    \[
    n \xrightarrow{f_1} n' \setminus (i' \setminus i) + j \xrightarrow{[g_1,c_2]} \mathcal{L}^{\bot} \xleftarrow{[g_2,c_1]} k' \setminus (j' \setminus j) + i \xleftarrow{f_2} k
    \]
\end{enumerate}
where $f_i$ and $g_i$ are defined as follows.  
 %because the image of $m$ is a convex down-closed e-hypergraph, 
The strictly internal interface $i' \setminus i$ of $\mathcal L$ is mapped to the internal interface $n'$ of $\mathcal G$,  since $\mathcal L$ is a down-closed subgraph of $\mathcal G$,  inducing an identification of $i' \setminus i$ in  $n'$. 
Then map $f_1$ is given by $g_{ext}$ with its codomain restricted to $n' \setminus (i' \setminus i)$.\footnote{Noting the image of $g_{ext}$ indeed lies within $n' \setminus (i' \setminus i)$}  The map $g_1$ is derived from the restriction of $g_{int}$ to type $n' \setminus (i' \setminus i) \to \mathcal G$ by further observing that $\mathcal L^\bot$ can be identified within $\mathcal G$ --- and in particular has the internal interface $n'$ of $\mathcal G$ minus $i' \setminus i$.  The maps $f_2$ and $g_2$ are defined similarly. 
%$i' \setminus i$, which is a strict internal interface of $\mathcal{L}$, is necessarily mapped to vertices in the strict internal interface of $\mathcal{G}$ which are also in the image of $n'$ and therefore we can construct a 
%morphism $f' : i' \setminus i \to n'$ by first following $m$ and then composing with the partial inverse of $g_{int}$ which exists because it is mono. 
%Then we define $n' \setminus (i' \setminus i)$ as the pushout complement in the diagram below
%
%\[
%    \tikzfig{combinatorial_semantics/pushout_difference}
%\]
%and $g_1$ is obtained by composition $g';g_{int}$ and then composing with the inverse of $l$.
%Similarly for $g_2$.
% CHRIS: TRY TO DEFINE G1, G2 AS SIMPLY AS POSSIBLE, AND POSSIBLY ADD AN INFORMAL EXPLANATION BELOW THE DEFINITION. 
% Because $\mathcal{L}^{\bot}$ is essentially $\mathcal{G}$ with $\mathcal{L}$ removed, apart from the part that has a pre-image in $i + j$, there is an obvious morphism from $l \to \mathcal{L}^{\bot}$ which is $g_1 = g_{int} |_{D}$ where
% \[
%     D = \{ v \in n' \text{ such that } \not \exists u \in (i' \setminus i) ~ . ~ g_{int}(v) = f'_{int};f(u) \}
% \]
% Similarly for $r \to \mathcal{L}^{\bot}$, which we denote as $g_2$.
% Then there exists a morphism $n \to l$, because there is a morphism $n \to n'$ and the external interfaces of $n'$ and $l$ are the same.
% That is, it is a morphism $f: n \to l$ such that $d_1 = f;g_1$.
% Similarly for $k \to r$.
\end{definition}

Intuitively the above definition means that when the occurrence of $\mathcal{L}$ within $\mathcal{G}$ is top-level then $\mathcal{L}$'s external interfaces become the part of $\mathcal{L}^{\bot}$'s external interfaces and when the occurrence is nested, those interfaces become the part of $\mathcal{L}^{\bot}$'s strict internal interfaces.
In both cases when removing the occurrence of $\mathcal{L}$ from $\mathcal{G}$, $\mathcal{L}$'s strict internal interfaces are removed from strict internal interfaces of $\mathcal{G}$ to form internal interfaces of the corresponding MDA cospans above.
The morphisms are then constructed correspondingly to require that inputs (respectively, outputs) of $\mathcal{R}$ are glued to the outputs (respectively, inputs) of $\mathcal{L}^{\bot}$.

@@

Note that,  in the above definition,  the extended cospan 
    \[
    n \xrightarrow{} n' \setminus (i' \setminus i) + j \xrightarrow{[g_1,c_2]} \mathcal{L}^{\bot} \xleftarrow{[g_2,c_1]} k' \setminus (j' \setminus j) + i \xleftarrow{} k
    \]
 is not necessarily well-typed.
This is because its input internal interface may contain vertices that previously were a part of the output internal interface and vice versa.
\begin{remark}
The boundary complement conditions, in particular, prohibit finding a match for $\mathcal L$ in $\mathcal G$,  below.  
\[
	\mathcal{L} = \scalebox{0.8}{\tikzfig{../figures/combinatorial_semantics/f_times_g}} \qquad \mathcal{G} = \scalebox{0.6}{\tikzfig{../figures/combinatorial_semantics/f_plus_g_inline}}
\]
\end{remark}

\begin{proposition}
\label{prop:boundary_unique}
    The boundary complement in~\ref{def:boundary_new} when exists is unique.
\end{proposition}
